### 2020-10-06 paper.Rmd
storing data to save time during rendering

source("misc-functions.R")

@@

For the purpose of problem formalization, I adopt the notation Steinley 
(2006) provided in his thorough synthesis of a half-century literature 
on k-means. K-means clustering is used to partition two-way, two-mode 
data (that is, $N$ elements each having $P$ attributes) into $K$ groups 
$(C_1., \ldots, C_K)$ where $C_k$ is the set of $n_k$ objects in group 
$k$. The partitioning is subject to the requirements that each element 
is assigned to exactly one group, formalized by the two constraints

@@

# Generate data and all partitions

set.seed(2207) # 2207

N <- 14
K <- 2
partitions <- generate_partitions(N, K)
features <- rnorm(N, 15, 3)

@@

df <- data.frame(
  diff_mean = all_objectives_mean,
  diff_var  = all_objectives_var
)

@@

A motivating example illustrates that aiming to simultaneously optimize 
similarity of means and variances at the same time is not only 
desirable, but may also be feasible. I created 14 data points from a 
univariate normal distribution ($M = 15$, $SD = 3$) and generated all `r 
length(partitions)` possible ways to partition the 14 data points into 
$K = 2$ equal-sized groups. I computed the difference in the two means 
and two variances for each partition. This simple difference metric was 
chosen for purposes of interpretability; in fact, there is a a 1-to-1 
correspondence between these differences and $SSE$ and 
$SSE_{\mathrm{VAR}}$, respectively (Figure 1).[^onlyonegroup] Figure 2 
illustrates similarity in means plotted against similarity in variances 
across all partitions. Ideally, a partition is chosen that minimizes 
the difference with regard to means and variances (see the triangular 
symbol). However, optimal k-means anticlustering would create two groups 
that are not particularly similar with regard to their variance and 
chose the partition that is represented by the squared symbol. The 
diamond symbol represents that partition that where the two groups have 
the least difference in variances. The black symbols illustrate the 
other members of the optimal pareto set.

@@

# Plot relationship between objectives
layout(matrix(c(1,2,3,3), 2, 2, byrow = TRUE), heights = c(1, 1.4))
plot(
  all_objectives_mean, 
  all_objectives_kMeans, 
  xlab = "mean(f1) - mean(f2)",
  ylab = "SSE",
  pch = 4, 
  cex = .5,
  col = "darkgrey"
)
legend("topright", legend = "A", cex = 1.3, bty = "n")
plot(
  all_objectives_var, 
  all_objectives_kVar, 
  xlab = "var(f1) - var(f2)",
  ylab = "SSE_var",
  pch = 4, 
  cex = .5,
  col = "darkgrey"
)
legend("topright", legend = "B", cex = 1.3, bty = "n")

@@

# Plot SSE vs SSE(VAR)
plot(
  df, pch = 4, col = "darkgrey", 
  las = 1, cex = .5,
  xlab = "mean(f1) - mean(f2)",
  ylab = "var(f1) - var(f2)"
)
legend("topright", legend = "C", cex = 1.3, bty = "n")

# Illustrate Pareto set
pareto_pts <- pareto_set(df, sense = "min")
points(df[pareto_pts, ], pch = 19)

# Illustrate best partition wrt SSE (difference in means)
points(
  all_objectives_mean[which.min(all_objectives_mean)],
  all_objectives_var[which.min(all_objectives_mean)],
  cex = 1.2, pch = 22, bg = "white", lwd = 1.7
)

df$sum <- rowSums(df)

# Illustrate best partition wrt SSE(VAR) (i.e., difference in variance)
points(
  all_objectives_mean[which.min(df$sum)],
  all_objectives_var[which.min(df$sum)],
  cex = 1.2, pch = 24, bg = "white", lwd = 1.7
)

# Illustrate best partition wrt simililarity combined similarity / variance
points(
  all_objectives_mean[which.min(df$diff_var)],
  all_objectives_var[which.min(df$diff_var)],
  cex = 1.2, pch = 23, bg = "white", lwd = 1.7
)
```

>>>>

source("misc-functions.R")
set.seed(2207) # 2207

@@

For the purpose of problem formalization, I adopt the notation Steinley 
(2006) provided in his comprehensive synthesis of the literature 
on k-means. K-means clustering is used to partition two-way, two-mode 
data (that is, $N$ elements each having $P$ attributes) into $K$ groups 
$(C_1., \ldots, C_K)$ where $C_k$ is the set of $n_k$ objects in group 
$k$. The partitioning is subject to the requirements that each element 
is assigned to exactly one group, formalized by the two constraints

@@

# First rendering is slow because the data needs to be generated
path <- "partition_objectives.csv"

if (!file.exists(path)) {
  # Generate data and all partitions
  N <- 14
  K <- 2
  partitions <- generate_partitions(N, K)
  features <- rnorm(N, 15, 3)

@@

  df <- data.frame(
    diff_mean = all_objectives_mean,
    diff_var  = all_objectives_var,
    kmeans = all_objectives_kMeans,
    kvar = all_objectives_kVar
  )

  write.table(
    df, path, row.names = FALSE, sep = ","
  )
} else {
  df <- read.csv(path)
}

@@

A motivating example illustrates that aiming to simultaneously optimize 
similarity of means and variances at the same time is not only 
desirable, but may also be feasible. I created 14 data points from a 
univariate normal distribution ($M = 15$, $SD = 3$) and generated all `r 
n_partitions(14, 2)` possible ways to partition the 14 data points into 
$K = 2$ equal-sized groups. I computed the difference in the two means 
and two variances for each partition. This simple difference metric was 
chosen for purposes of interpretability; in fact, there is a a 1-to-1 
correspondence between these differences and $SSE$ and 
$SSE_{\mathrm{VAR}}$, respectively (Figure 1).[^onlyonegroup] Figure 2 
illustrates similarity in means plotted against similarity in variances 
across all partitions. Ideally, a partition is chosen that minimizes 
the difference with regard to means and variances (see the triangular 
symbol). However, optimal k-means anticlustering would create two groups 
that are not particularly similar with regard to their variance and 
chose the partition that is represented by the squared symbol. The 
diamond symbol represents that partition that where the two groups have 
the least difference in variances. The black symbols illustrate the 
other members of the optimal pareto set.

@@

# Plot relationship between objectives
layout(matrix(c(1,2,3,3), 2, 2, byrow = TRUE), heights = c(1, 1.4))
plot(
  df[, c("diff_mean", "kmeans")],
  xlab = "mean(f1) - mean(f2)",
  ylab = "SSE",
  pch = 4, 
  cex = .5,
  col = "darkgrey"
)
legend("topright", legend = "A", cex = 1.3, bty = "n")
plot(
  df[, c("diff_var", "kvar")],
  xlab = "var(f1) - var(f2)",
  ylab = "SSE_var",
  pch = 4, 
  cex = .5,
  col = "darkgrey"
)
legend("topright", legend = "B", cex = 1.3, bty = "n")

@@

# Plot SSE vs SSE(VAR)
plot(
  df[, c("kmeans", "kvar")], 
  pch = 4, col = "darkgrey", 
  las = 1, cex = .5,
  xlab = "kmeans",
  ylab = "kvar"
)
legend("bottomleft", legend = "C", cex = 1.3, bty = "n")

# Illustrate Pareto set
#pareto_pts <- pareto_set(df[, c("kmeans", "kvar")])
#points(df[pareto_pts, c("kmeans", "kvar")], pch = 19)

# Illustrate best partition wrt difference in means
points(
  df$kmeans[which.max(df$kmeans)],
  df$kvar[which.max(df$kmeans)],
  cex = 1.2, pch = 22, bg = "white", lwd = 1.7
)

# Illustrate best partition wrt difference in variance
points(
  df$kmeans[which.max(df$kvar)],
  df$kvar[which.max(df$kvar)],
  cex = 1.2, pch = 24, bg = "white", lwd = 1.7
)

df$kplus <- rowSums(df[, c("kmeans", "kvar")])

# Illustrate best partition wrt simililarity combined similarity / variance
points(
  df$kmeans[which.max(df$kplus)],
  df$kvar[which.max(df$kplus)],
  cex = 1.2, pch = 23, bg = "white", lwd = 1.7
)
```
